A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-04 tex/codatadp.tex
Improve termination proof.

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Singleton}
\(
\inferrule{
  t, u \in \tEc
  \quad
  t, u : \tau,
  \tau \in \Codata,
  \mid \tau \mid = 1
}{
  \Ec := \Ec, t \teq u
}
\)
\\[3.7ex]
\rn{Split} 
\(
\inferrule{
  s^i_j( t ) \in \tEc \quad \text{ or } \quad t : \tau, \tau \in \Data, \tau \text{ finite }
}{
  \Ec := \Ec, t \teq \const{C}_1( s^1_1( t ), \ldots, s^{a_1}_1( t ) ) \ror \ldots \ror \Ec := \Ec, t \teq \const{C}_n( s^1_n( t ), \ldots, s^{a_n}_n( t ) ) 
}
\)
\end{tabular}
\caption{Rules for splitting.  
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

@@

Finally, if neither a rule from Figure~\ref{fig:cc-rules} or Figure~\ref{fig:ab-rules} applies to $\Ec$, 
the rules from Figure~\ref{fig:split-rule} are applied.
The rule \rn{Singleton} handles the degenerate case when there are two terms from $\tEc$ having a singleton codatatype type,
in which case we can infer they are equal.
Otherwise, in the case there is a selector term applied to $t$, or if $t$ has a finite datatype type, 
then we require the equivalence class of $t$ to contain a constructor term, which is enforced by the rule \rn{Split}.

@@

\begin{lemma}[Termination]
All derivation trees are finite.
\end{lemma}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $S^0$ be the set of terms that are arguments of a selector application in $\tEc$.
For $i \geq 0$, let $S^{i+1}$ $=$ $S_i \cup \{ s( t ) \mid t \in S^i, t : \tau, \tau \in \Data, \tau \in  \text{ finite}, s \in \Sel^\tau  \}$.
Let $S^\ast$ be the fixed point of this sequence, which is guarenteed to be a finite set 
since $S^i \setminus S^0$ only includes selectors applied to terms having finite datatype type.
Now, let $\tcEc$ be the set of subterms of $\Ec \cup \{ C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) ) \mid t \in S^\ast, t : \tau, C_i \in \Ctr^\tau \}$.
In a derivation tree with root node $\Ec$, 
it can be shown by case analysis on the rules of the calculus that each non-root node is 
the result of adding an equality between two terms from $\tcEc$ to its parent node.
Thus, the depth of a branch in a derivation tree with root node $\Ec$ is at most $\mid \tcEc \mid^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

\begin{lemma}[Refutation Soundness]
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Unify$_2$}, or \rn{Cyclic},
then $\Ec$ is unsatisfiable in $\thD$.
\rem{Elaborate on cyclic?}
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since all models of $\thD$ entail exactly one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is unsatisfiable in $\thD$.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, \rn{Unify$_1$}, \rn{Bisimilar}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
\rem{Elaborate on Bisimilar?}
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is unsatisfiable in $\thD$, 
and thus $\Ec$ is unsatisfiable in $\thD$.
\qed
\end{proof}

>>>>

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Singleton}
\(
\inferrule{
  t, u : \tau \in \tEc
  \quad
  \tau \in \Codata,
  \mid \tau \mid = 1
}{
  \Ec := \Ec, t \teq u
}
\)
\\[3.7ex]
\rn{Split} 
\(
\inferrule{
  t : \tau \in \tEc 
  \quad 
  ( s( t ) \in \tEc, s \in \Sel^\tau ) 
  \text{ or } 
  ( \tau \in \Data, \tau \text{ finite } )
}{
  \Ec := \Ec, t \teq \const{C}_1( s^1_1( t ), \ldots, s^{a_1}_1( t ) ) \ror \ldots \ror \Ec := \Ec, t \teq \const{C}_n( s^1_n( t ), \ldots, s^{a_n}_n( t ) ) 
}
\)
\end{tabular}
\caption{Rules for splitting.  
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

@@

Finally, if neither a rule from Figure~\ref{fig:cc-rules} or Figure~\ref{fig:ab-rules} applies to $\Ec$, 
the rules from Figure~\ref{fig:split-rule} are applied.
The rule \rn{Singleton} handles the degenerate case when there are two terms from $\tEc$ having a singleton codatatype type,
in which case they must be equivalent.
Otherwise, in the case there is a selector term applied to $t$, or if $t$ has a finite datatype type, 
then the equivalence class of $t$ must contain an application of one of the constructors of $\tau$ (where $\Ctr^\tau = \{ C_1, \ldots C_n \}$), 
which is enforced by the rule \rn{Split}.

@@

\begin{lemma}[Termination]
All derivation trees are finite.
\end{lemma}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $S$ be the set of terms occurring as the argument of a selector in $\tEc$.
For each term $t \in \tEc$ of finite datatype type $\tau$,
let $S^0_t$ be the set $\{ t \}$,
and for $i \geq 0$ let $S^{i+1}_t =$ $S^i_t \cup \{ s( u ) \mid u : \upsilon \in S^i, \upsilon \in \Data, \upsilon \text{ finite}, s \in \Sel^{\upsilon}  \}$.
Let $S^\ast_t$ be the fixed point of this sequence for all $t$, which we know is finite since the values of $\tau$ are of finite size.
Let $S^\ast$ be the union of $S^\ast_t$ for all such $t$,
and let $\tcEc$ be the set of subterms of $\Ec \cup \{ C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) ) \mid t : \tau \in S \cup S^\ast, C_i \in \Ctr^\tau \}$.
In a derivation tree with root node $\Ec$, 
it can be shown by case analysis on the rules of the calculus that each non-root node $\Ec'$ is such that 
$\mathcal{T}(\Ec') \subseteq \tcEc $, and moreover thus is the result of adding an equality between two terms from $\tcEc$ to its parent node.
Thus, the depth of a branch in a derivation tree with root node $\Ec$ is at most $\mid \tcEc \mid^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

\begin{lemma}[Refutation Soundness]
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Unify$_2$}, or \rn{Cyclic},
then $\Ec$ is unsatisfiable in $\thD$.
\rem{Elaborate on cyclic?}
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since all models of $\thD$ entail (exactly) one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is unsatisfiable in $\thD$.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, \rn{Unify$_1$}, \rn{Bisimilar}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
\rem{Elaborate on Bisimilar?}
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is unsatisfiable in $\thD$, 
and thus $\Ec$ is unsatisfiable in $\thD$.
\qed
\end{proof}
